---
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year.

## Import Library

```{r library, message=FALSE, warning=FALSE}
# install.packages("tidyverse")
library(tidyverse)
library(lubridate)
```

## Load Data

-   `read_csv()` reads comma delimited files

```{r load}
df = read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
head(df)
```

## Tidy and Transform Data

```{r}
df_2 = df %>% select(INCIDENT_KEY, 
                   OCCUR_DATE,
                   OCCUR_TIME,
                   BORO, 
                   STATISTICAL_MURDER_FLAG,
                   PERP_AGE_GROUP,
                   PERP_SEX,
                   PERP_RACE,
                   VIC_AGE_GROUP,
                   VIC_SEX,
                   VIC_RACE,
                   Latitude,
                   Longitude)

# Return the column name along with the missing values
lapply(df_2, function(x) sum(is.na(x)))
```

Understanding the reasons why data are missing is important for handling the remaining data correctly. There's a fair amount of unidentifiable data on perpetrators (age, race, or sex.) Those cases are possibly still active and ongoing investigation. In fear of missing meaningful information, I handle this group of missing data by calling them as another group of "Unknown".

Key observations on data type conversion are:

-   **INCIDENT_KEY** should be treated as a string.
-   **BORO** should be treated as a factor.
-   **PERP_AGE_GROUP** should be treated as a factor.
-   **PERP_SEX** should be treated as a factor.
-   **PERP_RACE** should be treated as a factor.
-   **VIC_AGE_GROUP** should be treated as a factor.
-   **VIC_SEX** should be treated as a factor.
-   **VIC_RACE** should be treated as a factor.

```{r}
# Tidy and transform data
df_2 = df_2 %>% 
  replace_na(list(PERP_AGE_GROUP = "Unknown", PERP_SEX = "Unknown", PERP_RACE = "Unknown"))

# Remove extreme values in data
df_2 = subset(df_2, PERP_AGE_GROUP!="1020" & PERP_AGE_GROUP!="224" & PERP_AGE_GROUP!="940")

df_2$PERP_AGE_GROUP = recode(df_2$PERP_AGE_GROUP, UNKNOWN = "Unknown")
df_2$PERP_SEX = recode(df_2$PERP_SEX, U = "Unknown")
df_2$PERP_RACE = recode(df_2$PERP_RACE, UNKNOWN = "Unknown")
df_2$VIC_SEX   = recode(df_2$VIC_SEX, U = "Unknown")
df_2$VIC_RACE   = recode(df_2$VIC_RACE, UNKNOWN = "Unknown")
df_2$INCIDENT_KEY = as.character(df_2$INCIDENT_KEY)
df_2$BORO = as.factor(df_2$BORO)
df_2$PERP_AGE_GROUP = as.factor(df_2$PERP_AGE_GROUP)
df_2$PERP_SEX = as.factor(df_2$PERP_SEX)
df_2$PERP_RACE = as.factor(df_2$PERP_RACE)
df_2$VIC_AGE_GROUP = as.factor(df_2$VIC_AGE_GROUP)
df_2$VIC_SEX = as.factor(df_2$VIC_SEX)
df_2$VIC_RACE = as.factor(df_2$VIC_RACE)

# Return summary statistics
summary(df_2)
```

## Visualizations and Analysis

Which part of New York has the most number of incidents? Of those incidents, how many are murder cases?

Brooklyn is the first in terms of the number of incidents, followed by Bronx and Queens respectively.

```{r}
g <- ggplot(df_2, aes(x = BORO)) +
  geom_bar() +
  labs(title = "Boroughs of New York City",
       x = "Boroughs of New York City",
       y = "Count of Incidents") +
  theme_minimal()
g
```

```{r}
table(df_2$BORO, df_2$STATISTICAL_MURDER_FLAG)
```

2.  Which day and time should people in New York be cautious of falling into victims of crime?

-   Weekends in NYC have the most chances of incidents.
-   Historically occurs in the evening and night time.

```{r}
df_2$OCCUR_DAY = mdy(df_2$OCCUR_DATE)
df_2$OCCUR_DAY = wday(df_2$OCCUR_DAY, label = TRUE)
df_2$OCCUR_HOUR = hour(hms(as.character(df_2$OCCUR_TIME)))

df_3 = df_2 %>%
  group_by(OCCUR_DAY) %>%
  count()

df_4 = df_2 %>%
  group_by(OCCUR_HOUR) %>%
  count()
```

```{r}
g <- ggplot(df_3, aes(x = OCCUR_DAY, y = n)) +
  geom_col() +
  labs(title = "Which day should people in New York be cautious of incidents?",
       x = "Incident Occurence Day",
       y = "Count of Incidents") +
  theme_minimal()
g
```

```{r}
g <- ggplot(df_4, aes(x = OCCUR_HOUR, y = n)) +
  geom_line() +
  labs(title = "Which time should people in New York be cautious of incidents?",
       x = "Incident Occurence Hour",
       y = "Count of Incidents") +
  theme_minimal()
g
```

3.  The Profile of Perpetrators and Victims

-   There's a significant number of incidents in the age group of 25-44 and 18-24.
-   African American and White Hispanic stood out in the number of incidents in Boroughs of New York City.
-   There are significantly more incidents with Male sex than those of Female sex.

```{r}
table(df_2$PERP_AGE_GROUP, df_2$VIC_AGE_GROUP)
```

```{r}
table(df_2$PERP_SEX, df_2$VIC_SEX)
```

```{r}
table(df_2$PERP_RACE, df_2$VIC_RACE)
```

## Step 4: Identifying Bias

This analysis provides valuable insights into the patterns and trends of shooting incidents in New York City over the years. Key findings include:

```         
Geographical Distribution: Brooklyn has the highest number of shooting incidents, followed by the Bronx and Queens.
  Factors like socioeconomic status, neighborhood characteristics, or law enforcement intensity are not included but may influence shooting incidents.

Temporal Patterns: Incidents peak during weekends and in the evening hours.

Demographics: Most incidents involve male perpetrators and victims, with significant representation in the 18-44 age groups.

  By filtering out extreme age groups, we might be excluding legitimate cases, potentially biasing the age-related analysis.
```

In addition, there may be reporting bias involved by the NYPD, which underreporting or inconsistencies in data entry can impact the accuracy of the above-model. Some of the bias also involves being labeled as 'Unknown' which in this case, is impacting the analysis.
